---
title: "Model Training and Evaluation Report"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    code-summary: "Mostra codice"
    df-print: paged
    page-layout: full
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---


```{r, echo=FALSE}
library(tidyverse)
library(tidymodels)
library(ranger)
library(gt)
library(vip)
library(patchwork)
# library(probably)
library(bundle)

source("metrics_report.R")
source("helpers_report.R")


```


# Key decisions made in the modeling process

- **Dataset composition:** combined data from 3 surveys (SN217, SN267, AVQ 2021) using class proportions observed in the surveys → N = 400, O = 200, S = 125  
- **Imputation:** no imputation applied (preliminary tests showed no improvement with imputation)  
- **Algorithm selection:** Random Forest with hyperparameter tuning (chosen after comparing with XGBoost with and without tuning).

# Indicators

```{r}
# Data import & preparation

# Import dataset
data <- read_csv("train_data.csv")


# Balanced sample
set.seed(123)

data_N <- data %>%
  filter(type == "N") %>%
  sample_n(400)

data_O <- data %>%
  filter(type == "O") %>%
  sample_n(200)

data_S <- data %>%
  filter(type == "S")

data <- bind_rows(data_N, data_O, data_S)


# Split into training (80%) and test (20%) sets, stratified by 'type'
set.seed(123)
data_split <- initial_split(data, prop = 0.8, strata = type)
train_data <- training(data_split)
test_data <- testing(data_split)

# Fix consistent class level ordering for the whole script
class_order <- c("N", "O", "S")


indicators <- train_data %>%
  select(-dataset, -variable, -type) %>%
  colnames()

# Create tibble

df_indicators <- tibble(indicator = indicators)

df_indicators %>%
  gt() %>%
  tab_header(
    title = paste0("List of Used Indicators (n = ", nrow(df_indicators), ")")
  ) %>%
  cols_label(
    indicator = ""
  )


# --- NA count per indicator --------------------------------------------------

na_table <- train_data %>%
  select(all_of(indicators)) %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "indicator", values_to = "na_count") %>%
  arrange(desc(na_count))

na_table %>%
  gt() %>%
  tab_header(
    title = "Number of Missing Values per Indicator"
  ) %>%
  cols_label(
    indicator = "Indicator",
    na_count = "NAs"
  )


numeric_data <- train_data %>%
  select(all_of(indicators)) %>%
  select(where(is.numeric))

# Compute correlation matrix
cor_matrix <- cor(numeric_data, use = "pairwise.complete.obs")

# Convert to long format for conditional formatting
cor_df <- as.data.frame(cor_matrix) %>%
  rownames_to_column(var = "indicator")

# Build GT table: format numbers + highlight abs(cor) > 0.6
gt(cor_df) %>%
  fmt_number(
    columns = -indicator,
    decimals = 1
  ) %>%
  data_color(
    columns = -indicator,
    fn = function(x) {
      ifelse(abs(x) > 0.6 & abs(x) != 1, "#ff9999", "white")
    }
  ) %>%
  tab_header(
    title = "Correlation Matrix of Indicators (1 decimal, |r| > 0.6 in red)"
  )


```

# Exploratory Analysis: Indicator Distributions
```{r}

data %>%
  select(-variable, -dataset, -type) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(value_plot = ifelse(is.na(value), NA_real_, value)) %>%
  ggplot(aes(x = variable, y = value_plot)) +
  geom_boxplot(outlier.colour = "red", outlier.shape = 1) +
  theme_minimal() +
  labs(
    title = "Boxplot delle metriche",
    x = "Variabile",
    y = "Valore",
    caption = "Outlier in rosso"
  ) +
  facet_wrap(~variable, scales = "free_y") +
  theme(axis.text.x = element_blank())

```

```{r}

# Define Preprocessing Recipes

# (A) Recipe with no imputation
rec_none <- recipe(type ~ ., data = train_data) %>%
  update_role(dataset, variable, new_role = "info")

# (B) Recipe with median imputation for all numeric predictors
# rec_median <- recipe(type ~ ., data = train_data) %>%
#  update_role(dataset, variable, new_role = "info") %>%
#  step_impute_median(all_numeric_predictors())

```


```{r}

# Model Specifications

# ---- Random Forest ----
rf_spec <- rand_forest(
  trees = tune(),
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine("ranger", importance = "permutation") %>%
  set_mode("classification")

```

# Hyperparameter Tuning for Random Forest

```{r}
# Create workflow combining recipe + model
rf_tune_wf <- workflow() %>%
  add_recipe(rec_none) %>% # change rec_median here for imputation
  add_model(rf_spec)

# 5-fold cross-validation (stratified)
set.seed(123)
cv_folds <- vfold_cv(train_data, v = 5, strata = type)

# Define regular grid for RF tuning
rf_grid <- grid_latin_hypercube(
  trees(range = c(100L, 500L)),
  mtry(range = c(2L, 7L)),
  min_n(range = c(2L, 10L)),
  size = 20
)

# Run grid search for RF
rf_tune_res <- tune_grid(
  rf_tune_wf,
  resamples = cv_folds,
  grid = rf_grid,
  metrics = metric_set(accuracy)
)


# Select best parameters based on accuracy
best_rf_params <- select_best(rf_tune_res, metric = "accuracy")
best_rf_params %>%
  gt() %>%
  tab_header(
    title = "Best Random Forest Hyperparameters"
  ) %>%
  fmt_number(
    columns = everything(),
    decimals = 2
  )

# Finalize model with best parameters
rf_final_spec <- finalize_model(rf_spec, best_rf_params)

rf_final_wf <- workflow() %>%
  add_recipe(rec_none) %>%
  add_model(rf_final_spec)

# Fit the tuned RF on the full training data
rf_final_fit <- rf_final_wf %>%
  fit(data = train_data)

# Save final Random Forest workflow with fitted model

rf_final_fit_bundled <- bundle(rf_final_fit)

usethis::use_data(rf_final_fit, overwrite = TRUE, internal = TRUE)

saveRDS(
  rf_final_fit_bundled,
  file = "rf_final_fit_bundle.rds"
)

```



# Evaluation

```{r, echo = FALSE}

rf_preds <- predict(rf_final_fit, test_data) %>%
  bind_cols(predict(rf_final_fit, test_data, type = "prob")) %>%
  bind_cols(test_data %>% select(type, dataset, variable)) %>%
  mutate(type = as.factor(type))

rf_cm <- conf_mat(rf_preds, truth = type, estimate = .pred_class)

cat("Confusion matrix for tuned Random Forest\n")

rf_cm


rf_cm_perc_col <- prop.table(rf_cm$table, margin = 2) * 100

cat("\nConfusion matrix for tuned Random Forest (percentages by column):\n")
print(round(rf_cm_perc_col, 1))

cm_mat <- rf_cm_perc_col

if (is.null(rownames(cm_mat))) {
  rownames(cm_mat) <- c("N", "O", "S")
}
if (is.null(colnames(cm_mat))) {
  colnames(cm_mat) <- c("N", "O", "S")
}

# Flatten matrix
rf_cm_flat <- expand.grid(
  Pred = rownames(cm_mat),
  Truth = colnames(cm_mat)
) %>%
  mutate(
    Freq = as.vector(cm_mat),
    name = paste0("truth", Truth, "_pred", Pred)
  ) %>%
  select(name, Freq) %>%
  pivot_wider(names_from = name, values_from = Freq) %>%
  mutate(model = "RF_tuned")


# Compute standard metrics
rf_metrics <- metric_set(accuracy, f_meas, kap)
rf_metrics_res <- rf_metrics(rf_preds, truth = type, estimate = .pred_class)

# Compute multiclass ROC AUC (macro average)
prob_cols <- paste0(".pred_", levels(rf_preds$type))
rf_roc_res <- roc_auc(
  rf_preds,
  truth = type,
  all_of(prob_cols),
  estimator = "macro"
)

# Combine all metrics
rf_metrics_res <- bind_rows(rf_metrics_res, rf_roc_res)


final_combined_wide <- rf_metrics_res %>%
  select(.metric, .estimate) %>%
  pivot_wider(names_from = .metric, values_from = .estimate)


final_combined_wide %>%
  gt() %>%
  tab_header(
    title = "Classification Model Performance Metrics"
  ) %>%
  fmt_number(
    columns = everything()[-1],
    decimals = 2
  ) %>%
  tab_options(
    table.font.names = "Arial",
    heading.title.font.size = 16,
    table.font.size = 12,
    table.width = pct(100)
  )


```


```{r,eval=FALSE}
# ---- XGBoost Evaluation ----

# Generate predictions (class + probabilities) on test data
xgb_preds <- predict(xgb_final_fit, test_data) %>%
  bind_cols(predict(xgb_final_fit, test_data, type = "prob")) %>%
  bind_cols(test_data %>% select(type, dataset, variable)) %>%
  mutate(type = as.factor(type))

# Compute confusion matrix (absolute counts)
xgb_cm <- conf_mat(xgb_preds, truth = type, estimate = .pred_class)
cat("\nConfusion matrix for tuned XGBoost (absolute counts):\n")
print(xgb_cm)

# Confusion matrix in percentages (per column)
xgb_cm_mat <- xgb_cm$table
xgb_cm_perc_col <- prop.table(xgb_cm_mat, margin = 2) * 100
cat("\nConfusion matrix for tuned XGBoost (percentages by column):\n")

xgb_cm_perc_col <- round(xgb_cm_perc_col, 1)
print(xgb_cm_perc_col)

# Prendi la matrice interna dalla conf_mat percentuale
cm_mat <- xgb_cm_perc_col # XGBoost con imputazione, percentuale per colonna

# Assicurati che righe e colonne abbiano nomi
if (is.null(rownames(cm_mat))) {
  rownames(cm_mat) <- c("N", "O", "S")
}
if (is.null(colnames(cm_mat))) {
  colnames(cm_mat) <- c("N", "O", "S")
}

# Trasforma in dataframe piatto
xgb_cm_flat <- expand.grid(
  Pred = rownames(cm_mat),
  Truth = colnames(cm_mat)
) %>%
  mutate(
    Freq = as.vector(cm_mat),
    name = paste0("truth", Truth, "_pred", Pred)
  ) %>%
  select(name, Freq) %>%
  pivot_wider(names_from = name, values_from = Freq) %>%
  mutate(model = "XGB_impute")


# Compute standard metrics (accuracy, F1-score, kappa)
xgb_metrics <- metric_set(accuracy, f_meas, kap)
xgb_metrics_res <- xgb_metrics(xgb_preds, truth = type, estimate = .pred_class)

# Compute multiclass ROC AUC (macro average)
prob_cols <- paste0(".pred_", levels(xgb_preds$type))
xgb_roc_res <- roc_auc(
  xgb_preds,
  truth = type,
  all_of(prob_cols),
  estimator = "macro"
)
xgb_roc_res <- xgb_roc_res %>%
  mutate(.estimator = "macro") %>%
  select(.metric, .estimate, .estimator)

xgb_metrics_res <- xgb_metrics_res %>%
  mutate(.estimator = NA) %>%
  bind_rows(xgb_roc_res)

# Combine all metrics into a single table
xgb_metrics_res <- bind_rows(xgb_metrics_res, xgb_roc_res)

# cat("\nPerformance metrics for tuned XGBoost:\n")
# print(xgb_metrics_res)

# ---- Summary comparison with Random Forest ----
final_results <- bind_rows(
  rf_metrics_res %>% mutate(model = "Random Forest"),
  xgb_metrics_res %>% mutate(model = "XGBoost")
)


# print(final_results)
```

```{r,eval=FALSE}

# old

# ---- Fit models with recipe without imputation ----

# ---- Random Forest ----

rf_none_wf <- workflow() %>%
  add_recipe(rec_none) %>%
  add_model(rf_final_spec)

rf_none_fit <- rf_none_wf %>%
  fit(data = train_data)

# Predictions

rf_none_preds <- predict(rf_none_fit, test_data) %>%
  bind_cols(predict(rf_none_fit, test_data, type = "prob")) %>%
  bind_cols(test_data %>% select(type, dataset, variable)) %>%
  mutate(type = factor(type, levels = class_order))

# Confusion matrix

rf_none_cm <- conf_mat(rf_none_preds, truth = type, estimate = .pred_class)
rf_none_cm_mat <- rf_none_cm$table
rf_none_cm_perc <- prop.table(rf_none_cm_mat, margin = 2) * 100

cat("Confusion matrix RF (no imputation):\n")
print(rf_none_cm)
cat("\nConfusion matrix RF (no imputation) % by column:\n")
print(round(rf_none_cm_perc, 1))

cm_mat <- rf_none_cm_perc

if (is.null(rownames(cm_mat))) {
  rownames(cm_mat) <- c("N", "O", "S")
}
if (is.null(colnames(cm_mat))) {
  colnames(cm_mat) <- c("N", "O", "S")
}

# Flatten matrix
rf_none_cm_flat <- expand.grid(
  Pred = rownames(cm_mat),
  Truth = colnames(cm_mat)
) %>%
  mutate(
    Freq = as.vector(cm_mat),
    name = paste0("truth", Truth, "_pred", Pred)
  ) %>%
  select(name, Freq) %>%
  pivot_wider(names_from = name, values_from = Freq) %>%
  mutate(model = "RF_no_impute") %>%
  mutate(across(where(is.numeric), ~ round(.x, 1)))


# Metrics

rf_none_metrics <- metric_set(accuracy, f_meas, kap)
rf_none_res <- rf_none_metrics(
  rf_none_preds,
  truth = type,
  estimate = .pred_class
)
rf_none_roc <- roc_auc(
  rf_none_preds,
  truth = type,
  all_of(paste0(".pred_", levels(rf_none_preds$type))),
  estimator = "macro"
)
rf_none_res <- bind_rows(rf_none_res, rf_none_roc) %>%
  mutate(model = "RF_no_impute")

# ---- XGBoost ----

xgb_none_wf <- workflow() %>%
  add_recipe(rec_none) %>%
  add_model(xgb_final_spec)

xgb_none_fit <- xgb_none_wf %>%
  fit(data = train_data)

# Predictions

xgb_none_preds <- predict(xgb_none_fit, test_data) %>%
  bind_cols(predict(xgb_none_fit, test_data, type = "prob")) %>%
  bind_cols(test_data %>% select(type, variable, dataset)) %>%
  mutate(type = factor(type, levels = class_order))

# Confusion matrix

xgb_none_cm <- conf_mat(xgb_none_preds, truth = type, estimate = .pred_class)
xgb_none_cm_mat <- xgb_none_cm$table
xgb_none_cm_perc <- prop.table(xgb_none_cm_mat, margin = 2) * 100

cat("\nConfusion matrix XGBoost (no imputation):\n")
print(xgb_none_cm)
cat("\nConfusion matrix XGBoost (no imputation) % by column:\n")
print(round(xgb_none_cm_perc, 1))

cm_mat <- xgb_none_cm_perc

if (is.null(rownames(cm_mat))) {
  rownames(cm_mat) <- c("N", "O", "S")
}
if (is.null(colnames(cm_mat))) {
  colnames(cm_mat) <- c("N", "O", "S")
}

# Trasforma in dataframe piatto
xgb_none_cm_flat <- expand.grid(
  Pred = rownames(cm_mat),
  Truth = colnames(cm_mat)
) %>%
  mutate(
    Freq = as.vector(cm_mat),
    name = paste0("truth", Truth, "_pred", Pred)
  ) %>%
  select(name, Freq) %>%
  pivot_wider(names_from = name, values_from = Freq) %>%
  mutate(model = "XGB_no_impute")

xgb_none_cm_flat <- xgb_none_cm_flat %>%
  mutate(across(where(is.numeric), round, 1))


# Metrics

xgb_none_metrics <- metric_set(accuracy, f_meas, kap)
xgb_none_res <- xgb_none_metrics(
  xgb_none_preds,
  truth = type,
  estimate = .pred_class
)
xgb_none_roc <- roc_auc(
  xgb_none_preds,
  truth = type,
  all_of(paste0(".pred_", levels(xgb_none_preds$type))),
  estimator = "macro"
)
xgb_none_res <- bind_rows(xgb_none_res, xgb_none_roc) %>%
  mutate(model = "XGB_no_impute")


rf_metrics_res <- rf_metrics_res %>% mutate(model = "RF_impute")
xgb_metrics_res <- xgb_metrics_res %>% mutate(model = "XGB_impute")


final_combined_results <- bind_rows(
  rf_metrics_res,
  xgb_metrics_res,
  rf_none_res,
  xgb_none_res
)

final_combined_results <- final_combined_results %>%
  filter(.metric != "roc_auc" | (.metric == "roc_auc" & .estimator == "macro"))


final_combined_wide <- final_combined_results %>%
  select(model, .metric, .estimate) %>%
  pivot_wider(names_from = .metric, values_from = .estimate)


final_combined_wide %>%
  gt() %>%
  tab_header(
    title = "Classification Model Performance Metrics"
  ) %>%
  fmt_number(
    columns = everything()[-1],
    decimals = 2
  ) %>%
  tab_options(
    table.font.names = "Arial",
    heading.title.font.size = 16,
    table.font.size = 12,
    table.width = pct(100)
  )


```




```{r}
rf_cm_flat %>%
  gt() %>%
  tab_header(title = "Summary of Confusion Matrices") %>%
  fmt_number(
    columns = everything()[-ncol(rf_cm_flat)],
    decimals = 1
  )
```

```{r, echo=FALSE}

# ROC curve

prob_cols <- paste0(".pred_", levels(rf_preds$type))

# Compute multiclass ROC curves correctly (one-vs-all)

# roc_rf <- roc_curve(rf_preds, truth = type, all_of(prob_cols)) %>%
#  mutate(model = "RF (impute)")

# roc_xgb <- roc_curve(xgb_preds, truth = type, all_of(prob_cols)) %>%
#  mutate(model = "XGB (impute)")

# roc_rf_none <- roc_curve(rf_none_preds, truth = type, all_of(prob_cols)) %>%
#  mutate(model = "RF (no impute)")

# roc_xgb_none <- roc_curve(xgb_none_preds, truth = type, all_of(prob_cols)) %>%
#  mutate(model = "XGB (no impute)")

# Combine for plotting

# roc_all <- bind_rows(roc_rf, roc_xgb, roc_rf_none, roc_xgb_none)

```


```{r, eval=FALSE}

# fig.width=10, fig.height=14, fig.align="center"
roc_all %>%
  ggplot(aes(x = 1 - specificity, y = sensitivity, color = model)) +
  geom_line(linewidth = 1.2) +
  facet_wrap(~.level, ncol = 1) +
  coord_equal() +
  theme_minimal() +
  theme(
    legend.position = "bottom"
  ) +
  labs(
    title = "Multiclass ROC Curves (tidymodels)",
    x = "1 - Specificity (False Positive Rate)",
    y = "Sensitivity (True Positive Rate)",
    color = "Model"
  )

```

# Variable importance scores
```{r, echo=FALSE}

rf_final_fit |>
  extract_fit_parsnip() |>
  vip(num_features = 10)

```

# Misclassification Analysis

```{r, echo=FALSE}

# Full predictions
rf_full_fit <- rf_final_wf %>%
  fit(data = data)

rf_preds_full <- predict(rf_full_fit, data) %>%
  bind_cols(predict(rf_full_fit, data, type = "prob")) %>%
  bind_cols(data %>% select(type, dataset, variable)) %>%
  mutate(type = factor(type, levels = class_order))

# Misclassified predictions
rf_mis_full <- rf_preds_full %>%
  filter(.pred_class != type) %>%
  select(type, .pred_class, dataset, variable, everything()) %>%
  arrange(dataset, variable)

cat("\n### Random Forest (no imputation) — FULL DATA\n")
rf_mis_full

```

```{r, echo=FALSE}

rf_flag <- rf_preds_full %>%
  select(variable, dataset, type, .pred_class) %>%
  mutate(correct_flag = if_else(type == .pred_class, "correct", "wrong"))


rf_flag_full <- rf_flag %>%
  left_join(data, by = c("dataset", "variable"))


pred_vars <- rf_flag_full %>%
  select(-dataset, -variable, -type.x, -.pred_class, -correct_flag) %>%
  select(where(is.numeric)) %>%
  names()


plot_correct_wrong <- function(data_subset, tipo) {
  plots <- lapply(pred_vars, function(var) {
    ggplot(
      data_subset,
      aes_string(x = "correct_flag", y = var, fill = "correct_flag")
    ) +
      geom_boxplot(alpha = 0.7) +
      coord_cartesian(
        ylim = quantile(data_subset[[var]], c(0.05, 0.95), na.rm = TRUE)
      ) +
      labs(
        title = paste("Distribution of", var, "for type", tipo),
        x = "Classification outcome",
        y = var
      ) +
      theme_minimal()
  })
  wrap_plots(plots, ncol = 2)
}


rf_nominal <- rf_flag_full %>% filter(type.x == "N")
rf_ordinal <- rf_flag_full %>% filter(type.x == "O")
rf_s <- rf_flag_full %>% filter(type.x == "S")

```



```{r rf_plot3, fig.width=14, fig.height=20}

cat("RF - Nominale\n")
plot_correct_wrong(rf_nominal, "N")

cat("RF - Ordinale\n")
plot_correct_wrong(rf_ordinal, "O")

cat("RF  - Tipo S\n")
plot_correct_wrong(rf_s, "S")

```


```{r}

rf_wrong <- rf_flag_full %>%
  filter(correct_flag == "wrong")


wrong_N <- rf_wrong %>% filter(type.x == "N")
wrong_O <- rf_wrong %>% filter(type.x == "O")
wrong_S <- rf_wrong %>% filter(type.x == "S")


data_sn217 <- read.csv(
  "/home/giovanna/codice/fossr-dc/_misc/SN217/SN217.csv",
  header = TRUE,
  sep = ",",
  stringsAsFactors = FALSE
)

data_sn267 <- read.csv(
  "/home/giovanna/codice/fossr-dc/_misc/SN267/SN267.csv",
  header = TRUE,
  sep = ",",
  stringsAsFactors = FALSE
)

data_avq21 <- readr::read_delim(
  "/home/giovanna/codice/fossr-dc/AVQ/2021/AVQ_Microdati_2021.txt",
  delim = "\t",
  col_types = readr::cols(.default = "n"),
  trim_ws = TRUE
)


original_datasets <- list(
  "SN217" = data_sn217,
  "SN267" = data_sn267,
  "AVQ21" = data_avq21
)

mis <- rf_flag_full %>%
  filter(correct_flag == "wrong") %>%
  distinct(dataset, variable, type.x, .pred_class)

```


```{r rf_plots2, eval=FALSE }

# fig.width=14, fig.height=10

rf_N <- rf_flag_full %>% filter(type.x == "N")

rf_N <- rf_N %>%
  mutate(group = if_else(correct_flag == "correct", "N→N", "N→O"))

na_percent_list <- list()

for (i in 1:nrow(rf_N)) {
  ds_name <- rf_N$dataset[i]
  varname <- rf_N$variable[i]
  group <- rf_N$group[i]

  df <- original_datasets[[ds_name]]
  if (!varname %in% names(df)) {
    next
  }

  values <- df[[varname]]
  na_percent <- sum(is.na(values)) / length(values) * 100

  na_percent_list[[length(na_percent_list) + 1]] <- data.frame(
    dataset = ds_name,
    variable = varname,
    group = group,
    na_percent = na_percent
  )
}

na_percent_df <- bind_rows(na_percent_list)

```

```{r}

pred_vars <- rf_flag_full %>%
  select(-dataset, -variable, -type.x, -.pred_class, -correct_flag) %>%
  select(where(is.numeric)) %>%
  names()

plot_by_class <- function(data, vars, class_col = "type.x") {
  plots <- lapply(vars, function(var) {
    # calcola limiti y
    vals <- data[[var]]
    y_min <- quantile(vals, 0.05, na.rm = TRUE)
    y_max <- quantile(vals, 0.95, na.rm = TRUE)

    ggplot(data, aes_string(x = class_col, y = var, fill = class_col)) +
      geom_boxplot(alpha = 0.7) +
      coord_cartesian(ylim = c(y_min, y_max)) +
      labs(
        title = paste("Distribution of", var, "by", class_col),
        x = class_col,
        y = var
      ) +
      theme_minimal() +
      theme(legend.position = "none")
  })

  wrap_plots(plots, ncol = 2)
}
```

## Distribution of Predictor Variables by True Class (N, O, S)

```{r rf_plots4, fig.width=14, fig.height=20}

plot_by_class(rf_flag_full, pred_vars, class_col = "type.x")

```

## Strong Correlations Among Predictor Variables (|r| > 0.6)
```{r}

cor_matrix <- data %>%
  select(all_of(pred_vars)) %>%
  cor(use = "pairwise.complete.obs")

cor_long <- as.data.frame(cor_matrix) %>%
  rownames_to_column("Var1") %>%
  pivot_longer(-Var1, names_to = "Var2", values_to = "Correlation")


cor_long <- as.data.frame(cor_matrix) %>%
  rownames_to_column("Var1") %>%
  pivot_longer(-Var1, names_to = "Var2", values_to = "Correlation") %>%
  filter(abs(Correlation) > 0.6 & abs(Correlation) < 0.99) %>%
  arrange(desc(abs(Correlation)))

cor_long %>%
  gt(rowname_col = "Var1") %>%
  tab_header(title = "Correlation Matrix ") %>%
  fmt_number(columns = "Correlation", decimals = 2) %>%
  data_color(
    columns = "Correlation",
    colors = col_numeric(
      palette = c("blue", "white", "red"),
      domain = c(-1, 1)
    )
  )

```

## Missing Values per Variable
```{r}

na_summary <- data %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "NA_count") %>%
  mutate(Total = nrow(data), NA_percent = NA_count / Total * 100) %>%
  arrange(desc(NA_count))


```



```{r,eval=FALSE}

# rf_plots, fig.width=14, fig.height=20
rf_flag_full %>%
  filter(type.x %in% c("N", "O")) %>%
  select(all_of(pred_vars), type.x) %>%
  pivot_longer(-type.x, names_to = "Variable", values_to = "Value") %>%
  ggplot(aes(x = type.x, y = Value, fill = type.x)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~Variable, scales = "free_y") +
  theme_minimal()

```


# Weighted Random Forest Classification

```{r}
set.seed(123)


class_weights <- train_data %>%
  count(type) %>%
  mutate(weight = sum(n) / (length(n) * n)) %>%
  deframe()


rf_weighted_spec <- rand_forest(
  trees = tune(),
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine(
    "ranger",
    importance = "permutation",
    class.weights = class_weights
  ) %>%
  set_mode("classification")


rf_weighted_wf <- workflow() %>%
  add_recipe(rec_none) %>%
  add_model(rf_weighted_spec)


rf_weighted_res <- tune_grid(
  rf_weighted_wf,
  resamples = cv_folds,
  grid = rf_grid,
  metrics = metric_set(bal_accuracy, f_meas, kap)
)


best_rf_weighted <- select_best(rf_weighted_res, metric = "bal_accuracy")

best_rf_weighted %>%
  gt() %>%
  tab_header(title = "Best Weighted Random Forest Parameters")


rf_weighted_final <- finalize_model(rf_weighted_spec, best_rf_weighted)

rf_weighted_final_wf <- workflow() %>%
  add_recipe(rec_none) %>%
  add_model(rf_weighted_final)

rf_weighted_fit <- rf_weighted_final_wf %>%
  fit(data = train_data)

# EVALUATION

rf_weighted_preds <- predict(rf_weighted_fit, test_data) %>%
  bind_cols(predict(rf_weighted_fit, test_data, type = "prob")) %>%
  bind_cols(test_data %>% select(type, dataset, variable)) %>%
  mutate(
    type = factor(type, levels = c("N", "O", "S")),
    .pred_class = factor(.pred_class, levels = c("N", "O", "S"))
  )

# ---- Balanced classification metrics ----
rf_weighted_metrics <- metric_set(
  accuracy,
  bal_accuracy,
  f_meas,
  recall,
  precision,
  kap
)(rf_weighted_preds, truth = type, estimate = .pred_class)

# ---- Multi-class ROC-AUC (probabilities) ----
rf_weighted_roc_auc <- rf_weighted_preds %>%
  roc_auc(truth = type, .pred_N, .pred_O, .pred_S)

# ---- Confusion Matrix ----
rf_weighted_confmat <- rf_weighted_preds %>%
  conf_mat(truth = type, estimate = .pred_class)

rf_weighted_cm_perc_col <- prop.table(rf_weighted_confmat$table, margin = 2) *
  100

cat("\nConfusion matrix for Weighted Random Forest (percentages by column):\n")
print(round(rf_weighted_cm_perc_col, 1))

rf_weighted_cm_table <- as.data.frame(rf_weighted_cm_perc_col) %>%
  rename(
    TrueClass = Truth,
    Predicted = Prediction,
    Percent = Freq
  ) %>%
  mutate(
    Percent = sprintf("%.1f%%", Percent)
  )

rf_weighted_cm_table
# ---- Display metrics ----
rf_weighted_metrics %>%
  gt() %>%
  tab_header(title = "Weighted Random Forest – Balanced Metrics")

rf_weighted_roc_auc %>%
  gt() %>%
  tab_header(title = "Weighted Random Forest – Multi-class ROC-AUC")

rf_weighted_confmat %>%
  autoplot(type = "heatmap") +
  ggtitle("Weighted Random Forest – Confusion Matrix")

saveRDS(rf_weighted_fit, "rf_weighted_final_fit.rds")

```
