---
title: "Typerclass: Model Training and Evaluation Report"
vignette: >
  %\VignetteIndexEntry{Typerclass: Model Training and Evaluation Report}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
execute: 
  cache: true
  echo: false
  warning: false
  message: false
---

```{r setup-libraries, cache=FALSE}
library(tidyverse)
library(tidymodels)
library(ranger)
library(gt)
library(vip)
library(patchwork)
library(bundle)
library(xgboost)
library(here)

knitr::opts_knit$set(root.dir = here::here())
```

```{r source-helpers, echo=FALSE}
source("./R/metrics.R")
source("./R/helpers.R")
```

## Introduction

This document presents the model training and evaluation process carried out
using Typerclass. The goal is to provide a clear overview of the methodological
choices and results of the modeling process.


## Key decisions in the modeling process

The model was trained following a few important choices regarding the data and
the algorithm.

### Dataset composition
Data from three surveys available at the Italian Istitute of Statistics (ISTAT)
were combined:

  - [Family, social subjects and life cycle](https://www.istat.it/en/microdata/multipurpose-survey-on-households-families-social-subjects-and-life-cycle-public-use-micro-stat-files/)
  - [Labour Force Survey](https://www.istat.it/en/microdata/labour-force-survey-cross-sectional-quarterly-data-2/)
  - [Aspects of daily life](https://www.istat.it/en/microdata/aspects-of-daily-life/)

The final dataset included 400 instances of class "Nominal", 200 of class
"Ordinal", and 125 of class "Scale" (mapped in the code as N, O, and S,
respectively).


## Indicators

The following indicators are used in the model, along with a brief description
of each:

- **n_unique_values** – Number of unique values in the variable (excluding
  missing values).  
- **std_dev** – Standard deviation; measures how spread out the values are.  
- **max_relative_frequency** – Proportion of the most frequent value relative to
  the total number of observations.  
- **norm_entropy** – Normalized entropy; measures how evenly the values are  
  distributed.
- **min_value** – Minimum observed value.  
- **max_value** – Maximum observed value.  
- **range_value** – Difference between the maximum and minimum values.
- **shannon_entropy** – Shannon entropy; a measure of uncertainty or information  
  content.
- **simpson_index** – Simpson diversity index; indicates how concentrated the
  values are.  
- **skewness_probs** – Skewness of the value-probability distribution; measures  
  asymmetry.
- **kurtosis_probs** – Excess kurtosis of the value distribution; indicates tail  
  heaviness.
- **dispersion_index** – Variance-to-mean ratio of value probabilities; measures  
  dispersion.
- **uniformity** – Shannon entropy normalized by `log(n_unique_values)`;
  measures distributional evenness.  
- **top2_ratio** – Sum of the probabilities of the two most frequent values.  
- **top3_ratio** – Sum of the probabilities of the three most frequent values.  

The table provides an overview of all indicators used in the model. Most
indicators have no missing values, ensuring reliable inputs for training. The
only exception is `dispersion_index`, which contains 128 missing values. The
`ranger` package can handle missing values by default; preliminary tests also
showed that imputation did not improve model performance.

```{r data-import-split}
# Data import & preparation

# Import dataset
data <- read_csv("./vignettes/model_data.csv")

# Balanced sample
set.seed(123)

data_N <- data %>%
  filter(type == "N") %>%
  sample_n(400)

data_O <- data %>%
  filter(type == "O") %>%
  sample_n(200)

data_S <- data %>%
  filter(type == "S")

data <- bind_rows(data_N, data_O, data_S)


# Split into training (80%) and test (20%) sets, stratified by 'type'
set.seed(123)
data_split <- initial_split(data, prop = 0.8, strata = type)
train_data <- training(data_split)
test_data <- testing(data_split)

# Fix consistent class level ordering for the whole script
class_order <- c("N", "O", "S")

indicators <- train_data %>%
  select(-dataset, -variable, -type) %>%
  colnames()

df_indicators <- tibble(indicator = indicators)


# --- NA count per indicator

na_table <- train_data %>%
  select(all_of(indicators)) %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "indicator", values_to = "na_count") %>%
  arrange(desc(na_count))

na_table %>%
  gt() %>%
  tab_header(
    title = "Overview of Indicators and Missing Values"
  ) %>%
  cols_label(
    indicator = "Indicator",
    na_count = "NAs"
  )
```


## Indicator Distributions

This section shows the distributions of all indicators, highlighting variability
and potential outliers. Several indicators display heavy tails and numerous
outliers, while others (e.g., proportion-based measures) are more tightly
concentrated; overall, the plots suggest substantial heterogeneity across
predictors.

To improve readability, a log transformation (log10 with +1) was
applied to a subset of indicators with very different scales or heavy tails. The
remaining panels are shown on the original scale so that comparably scaled
indicators can be interpreted directly.

```{r indicator-boxplots}

data %>%
  select(-variable, -dataset, -type) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(
    value_plot = ifelse(is.na(value), NA_real_, value),
    value_plot = case_when(
      variable %in%
        c(
          "max_value",
          "min_value",
          "range_value",
          "std_dev",
          "n_unique_values",
          "kurtosis_probs",
          "skewness_probs",
          "shannon_entropy"
        ) &
        !is.na(value_plot) &
        value_plot >= 0 ~ log10(value_plot + 1),
      TRUE ~ value_plot
    )
  ) %>%
  ggplot(aes(x = variable, y = value_plot)) +
  geom_boxplot(outlier.colour = "red", outlier.shape = 1) +
  theme_minimal() +
  labs(
    title = "Boxplot of Metrics (log10 for selected indicators)",
    x = "Variable",
    y = "Value (log10 for selected indicators)",
    caption = "Outliers in red"
  ) +
  facet_wrap(~variable, scales = "free_y", ncol = 3) +
  theme(axis.text.x = element_blank())

```


## Correlation matrix

The correlation matrix displays pairwise relationships between all model
indicators. Several predictors show high correlations, but they were retained in
the dataset because Random Forest — one of the algorithms selected for this
study — is robust to multicollinearity. In fact, including correlated indicators
can still improve model performance by providing additional predictive
information.

```{r correlation-matrix}

numeric_data <- train_data %>%
  select(all_of(indicators)) %>%
  select(where(is.numeric))

# Compute correlation matrix
cor_matrix <- cor(numeric_data, use = "pairwise.complete.obs")

# Convert to long format for conditional formatting
cor_df <- as.data.frame(cor_matrix) %>%
  rownames_to_column(var = "indicator")

# Build GT table: format numbers + highlight abs(cor) > 0.6
gt(cor_df) %>%
  fmt_number(
    columns = -indicator,
    decimals = 1
  ) %>%
  data_color(
    columns = -indicator,
    fn = function(x) {
      ifelse(abs(x) > 0.6 & abs(x) != 1, "#ff9999", "white")
    }
  ) %>%
  tab_header(
    title = "Correlation Matrix of Indicators"
  )
```


## Model Selection

### Preprocessing Recipe 

We tested a preprocessing approach using median imputation for all numeric
predictors. However, preliminary tests showed no improvement in model
performance, so we decided to proceed with a recipe without any imputation.

```{r preprocess-recipe}

# Define Preprocessing Recipes

# (A) Recipe with no imputation
rec_none <- recipe(type ~ ., data = train_data) %>%
  update_role(dataset, variable, new_role = "info")
```

Random Forest and XGBoost were both selected and evaluated with hyperparameter
tuning.

### Model Specification: Random Forest

```{r rf-spec}
# Model Specifications

# ---- Random Forest ----
rf_spec <- rand_forest(
  trees = tune(),
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine("ranger", importance = "permutation") %>%
  set_mode("classification")

```

### Hyperparameter Tuning for Random Forest

We selected the `mtry` range using the classic rule-of-thumb centered on
`sqrt(p)`, where `p` is the number of predictors, and expanded it by ±50% to
allow slightly simpler or more complex splits.

```{r rf-tuning}
# Create workflow combining recipe + model
rf_tune_wf <- workflow() %>%
  add_recipe(rec_none) %>% # change rec_median here for imputation
  add_model(rf_spec)

# 5-fold cross-validation (stratified)
set.seed(123)
cv_folds <- vfold_cv(train_data, v = 5, strata = type)

# Define regular grid for RF tuning
p <- ncol(train_data %>% select(-dataset, -variable, -type))
mtry_center <- floor(sqrt(p))
mtry_low <- max(1L, floor(mtry_center * 0.5))
mtry_high <- min(p, ceiling(mtry_center * 1.5))

rf_grid <- grid_space_filling(
  trees(range = c(300L, 1000L)),
  mtry(range = c(mtry_low, mtry_high)),
  min_n(range = c(2L, 20L)),
  size = 20
)

# Run grid search for RF
rf_tune_res <- tune_grid(
  rf_tune_wf,
  resamples = cv_folds,
  grid = rf_grid,
  metrics = metric_set(accuracy)
)


# Select best parameters based on accuracy
best_rf_params <- select_best(rf_tune_res, metric = "accuracy")
best_rf_params %>%
  gt() %>%
  tab_header(
    title = "Best Random Forest Hyperparameters"
  ) %>%
  fmt_number(
    columns = everything(),
    decimals = 2
  )

# Finalize model with best parameters
rf_final_spec <- finalize_model(rf_spec, best_rf_params)

rf_final_wf <- workflow() %>%
  add_recipe(rec_none) %>%
  add_model(rf_final_spec)

# Fit the tuned RF on the full training data
rf_final_fit <- rf_final_wf %>%
  fit(data = train_data)
```

### Model Specification: XGBoost

XGBoost was specified using a gradient-boosted tree model with tunable
hyperparameters. We used the `xgboost` engine and set the model to
classification mode to predict the three classes.

```{r xgb-spec}
# ---- XGBoost ----
xgb_spec <- boost_tree(
  trees = tune(),
  tree_depth = tune(),
  learn_rate = tune(),
  loss_reduction = tune(),
  min_n = tune(),
  sample_size = tune(),
  mtry = tune()
) %>%
  set_engine("xgboost") %>%
  set_mode("classification")
```

### Hyperparameter Tuning for XGBoost

We tuned XGBoost using the same cross-validation folds as Random Forest. The
`mtry` range follows the same rule-of-thumb (centered on `sqrt(p)`), while the
other hyperparameters use typical ranges for boosted trees.

```{r xgb-tuning}
# Create workflow combining recipe + model
xgb_tune_wf <- workflow() %>%
  add_recipe(rec_none) %>%
  add_model(xgb_spec)

# Define regular grid for XGBoost tuning
p <- ncol(train_data %>% select(-dataset, -variable, -type))
mtry_center <- floor(sqrt(p))
mtry_low <- max(1L, floor(mtry_center * 0.5))
mtry_high <- min(p, ceiling(mtry_center * 1.5))

xgb_params <- extract_parameter_set_dials(xgb_spec) %>%
  update(
    trees = trees(range = c(300L, 1000L)),
    tree_depth = tree_depth(range = c(2L, 10L)),
    learn_rate = learn_rate(range = c(0.01, 0.3)),
    loss_reduction = loss_reduction(range = c(0.0, 5.0)),
    min_n = min_n(range = c(2L, 20L)),
    sample_size = sample_prop(range = c(0.5, 1.0)),
    mtry = mtry(range = c(mtry_low, mtry_high))
  )

xgb_grid <- grid_space_filling(
  xgb_params,
  size = 20
)

# Run grid search for XGBoost
xgb_tune_res <- tune_grid(
  xgb_tune_wf,
  resamples = cv_folds,
  grid = xgb_grid,
  metrics = metric_set(accuracy)
)


# Select best parameters based on accuracy
best_xgb_params <- select_best(xgb_tune_res, metric = "accuracy")
best_xgb_params %>%
  gt() %>%
  tab_header(
    title = "Best XGBoost Hyperparameters"
  ) %>%
  fmt_number(
    columns = everything(),
    decimals = 3
  )

# Finalize model with best parameters
xgb_final_spec <- finalize_model(xgb_spec, best_xgb_params)

xgb_final_wf <- workflow() %>%
  add_recipe(rec_none) %>%
  add_model(xgb_final_spec)

# Fit the tuned XGBoost on the full training data
xgb_final_fit <- xgb_final_wf %>%
  fit(data = train_data)
```


## Evaluation

### Confusion matrix

We compare Random Forest and XGBoost on the same held-out test set. The confusion
matrices are normalized by row (true class), so values represent per-class recall.

```{r eval-confusion, echo = FALSE}

class_order <- c("N", "O", "S")

make_preds <- function(fit, data, model_name) {
  preds <- predict(fit, data) %>%
    bind_cols(predict(fit, data, type = "prob")) %>%
    bind_cols(data %>% select(type, dataset, variable)) %>%
    mutate(
      type = factor(type, levels = class_order),
      .pred_class = factor(.pred_class, levels = class_order),
      model = model_name
    )
  preds
}

cm_flat_from_preds <- function(preds) {
  cm <- conf_mat(preds, truth = type, estimate = .pred_class)
  cm_perc_row <- prop.table(cm$table, margin = 1) * 100

  cm_mat <- cm_perc_row
  rownames(cm_mat) <- class_order
  colnames(cm_mat) <- class_order

  expand.grid(
    Pred = rownames(cm_mat),
    Truth = colnames(cm_mat)
  ) %>%
    mutate(
      Freq = as.vector(cm_mat),
      name = paste0("truth", Truth, "_pred", Pred),
      model = unique(preds$model)
    ) %>%
    select(name, Freq, model) %>%
    pivot_wider(names_from = name, values_from = Freq)
}

rf_preds <- make_preds(rf_final_fit, test_data, "RF_tuned")
xgb_preds <- make_preds(xgb_final_fit, test_data, "XGB_tuned")

rf_cm_flat <- cm_flat_from_preds(rf_preds)
xgb_cm_flat <- cm_flat_from_preds(xgb_preds)

cm_compare <- bind_rows(rf_cm_flat, xgb_cm_flat)

rf_metrics <- metric_set(accuracy, bal_accuracy, f_meas, kap)
rf_metrics_res <- rf_metrics(
  rf_preds,
  truth = type,
  estimate = .pred_class,
  estimator = "macro"
)

xgb_metrics_res <- rf_metrics(
  xgb_preds,
  truth = type,
  estimate = .pred_class,
  estimator = "macro"
)

prob_cols <- paste0(".pred_", class_order)
rf_roc_res <- roc_auc(
  rf_preds,
  truth = type,
  all_of(prob_cols),
  estimator = "macro"
)
xgb_roc_res <- roc_auc(
  xgb_preds,
  truth = type,
  all_of(prob_cols),
  estimator = "macro"
)

metrics_compare <- bind_rows(
  rf_metrics_res %>% mutate(model = "RF_tuned"),
  rf_roc_res %>% mutate(model = "RF_tuned"),
  xgb_metrics_res %>% mutate(model = "XGB_tuned"),
  xgb_roc_res %>% mutate(model = "XGB_tuned")
)

```

```{r confusion-table}
cm_compare %>%
  gt() %>%
  tab_header(title = "Summary of Confusion Matrix (Per-Class Recall, %)") %>%
  cols_label(
    truthN_predN = "True N → Pred N",
    truthN_predO = "True N → Pred O",
    truthN_predS = "True N → Pred S",
    truthO_predN = "True O → Pred N",
    truthO_predO = "True O → Pred O",
    truthO_predS = "True O → Pred S",
    truthS_predN = "True S → Pred N",
    truthS_predO = "True S → Pred O",
    truthS_predS = "True S → Pred S",
    model = "Model"
  ) %>%
  fmt_number(
    columns = everything()[-ncol(cm_compare)],
    decimals = 1
  )
```

### Performance metrics

The table below compares Random Forest and XGBoost on the same test set using a
consistent set of metrics (accuracy, balanced accuracy, macro F1, Cohen’s
Kappa, and macro ROC AUC). Hyperparameters were selected using accuracy for
simplicity; we report additional metrics to assess class‑balanced performance.

```{r model-comparison-metrics, echo = FALSE}

metrics_table <- metrics_compare %>%
  select(model, .metric, .estimate) %>%
  pivot_wider(names_from = .metric, values_from = .estimate)

winner_row <- metrics_compare %>%
  group_by(.metric) %>%
  slice_max(.estimate, with_ties = FALSE) %>%
  ungroup() %>%
  select(.metric, model) %>%
  pivot_wider(names_from = .metric, values_from = model) %>%
  mutate(model = "Winner") %>%
  select(model, everything())

metrics_table_display <- metrics_table %>%
  mutate(across(-model, ~ sprintf("%.2f", .x))) %>%
  bind_rows(winner_row)

metric_winners <- metrics_compare %>%
  group_by(.metric) %>%
  slice_max(.estimate, with_ties = FALSE) %>%
  ungroup()

best_model <- metric_winners %>%
  count(model, name = "wins") %>%
  slice_max(wins, with_ties = FALSE) %>%
  pull(model)

metrics_table_display %>%
  gt() %>%
  tab_header(
    title = "Model Comparison Metrics"
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = model == "Winner")
  ) %>%
  tab_options(
    table.font.names = "Arial",
    heading.title.font.size = 16,
    table.font.size = 12,
    table.width = pct(100)
  )


```

Accuracy is the overall proportion of correct predictions, while balanced
accuracy averages recall across classes to reduce class-imbalance effects. Macro
F1 gives equal weight to each class, Cohen’s Kappa adjusts for chance agreement,
and macro ROC AUC summarizes discriminative ability across classes. Based on the
average of these metrics, the best overall model is **`r best_model`**.

### Misclassification Analysis

Misclassifications are inspected on the held-out test set to avoid optimistic
bias. The tables below summarize the misclassification patterns for each model.
```{r misclassification-prep, echo=FALSE}

misclassified <- bind_rows(
  rf_preds %>%
    filter(.pred_class != type) %>%
    mutate(model = "RF_tuned"),
  xgb_preds %>%
    filter(.pred_class != type) %>%
    mutate(model = "XGB_tuned")
) %>%
  select(model, type, .pred_class, dataset, variable, starts_with(".pred_")) %>%
  arrange(model, dataset, variable)

mis_counts <- misclassified %>%
  count(model, type, .pred_class, name = "n") %>%
  arrange(model, desc(n))

rf_mis_counts <- mis_counts %>%
  filter(model == "RF_tuned") %>%
  select(-model) %>%
  arrange(type, .pred_class)

xgb_mis_counts <- mis_counts %>%
  filter(model == "XGB_tuned") %>%
  select(-model) %>%
  arrange(type, .pred_class)

top_rf <- rf_mis_counts %>%
  slice_max(n, n = 1, with_ties = FALSE)
top_xgb <- xgb_mis_counts %>%
  slice_max(n, n = 1, with_ties = FALSE)

rf_total <- sum(rf_mis_counts$n)
xgb_total <- sum(xgb_mis_counts$n)

```

```{r misclassification-tables, echo=FALSE}
#| column: page-inset
#| layout-nrow: 1

rf_mis_counts %>%
  mutate(spacer = "") %>%
  gt() %>%
  tab_header(title = "Misclassification Summary (Test Set) — RF") %>%
  cols_label(
    type = "True class",
    .pred_class = "Predicted class",
    n = "Count",
    spacer = ""
  ) %>%
  cols_width(spacer ~ px(24)) %>%
  tab_style(
    style = cell_borders(sides = "all", color = "transparent"),
    locations = list(
      cells_body(columns = spacer),
      cells_column_labels(columns = spacer)
    )
  )

xgb_mis_counts %>%
  gt() %>%
  tab_header(title = "Misclassification Summary (Test Set) — XGBoost") %>%
  cols_label(
    type = "True class",
    .pred_class = "Predicted class",
    n = "Count"
  )
```

For RF, the most frequent error is `r top_rf$type[1]` →
`r top_rf$.pred_class[1]` (`r top_rf$n[1]` cases, `r rf_total` total
misclassifications). For XGBoost, the top confusion is `r top_xgb$type[1]` →
`r top_xgb$.pred_class[1]` (`r top_xgb$n[1]` cases, `r xgb_total` total). These
summaries highlight whether errors cluster between adjacent classes or are more
diffuse; fewer and more concentrated errors generally indicate a more reliable
model.

**Consistent with the overall performance metrics, Random Forest remains the
best-performing model and will be used for the final classifier.**

### Variable importance scores

This plot reports permutation importance for the Random Forest model. For each
indicator, its values are randomly permuted and the resulting drop in model
performance is measured; larger drops indicate more important predictors.
Because several indicators are correlated (as shown in the correlation section),
importance can be shared across related features, so the plot should be read as
a relative ranking rather than an absolute measure of effect.

```{r rf-variable-importance, echo=FALSE}

rf_final_fit |>
  extract_fit_parsnip() |>
  vip(num_features = 10)

```


### Distribution of Predictor Variables by True Class (N, O, S)

The final figure shows the distribution of predictor variables by their true
class (N, O, S) on the test set, highlighting how indicator values vary across
classes.

```{r predictor-distributions, fig.width=14, fig.height=20}

pred_vars <- test_data %>%
  select(-dataset, -variable, -type) %>%
  select(where(is.numeric)) %>%
  names()

plot_by_class <- function(data, vars, class_col = "type") {
  plots <- lapply(vars, function(var) {
    vals <- data[[var]]
    y_min <- quantile(vals, 0.05, na.rm = TRUE)
    y_max <- quantile(vals, 0.95, na.rm = TRUE)

    ggplot(data, aes_string(x = class_col, y = var, fill = class_col)) +
      geom_boxplot(alpha = 0.7) +
      coord_cartesian(ylim = c(y_min, y_max)) +
      labs(
        title = paste("Distribution of", var, "by", class_col),
        x = class_col,
        y = var
      ) +
      theme_minimal() +
      theme(legend.position = "none")
  })

  wrap_plots(plots, ncol = 3)
}

plot_by_class(test_data, pred_vars, class_col = "type")

```

```{r}
# Save final Random Forest workflow with fitted model
# Avoid writing during R CMD check (vignette build runs in a temp, read-only dir)
if (interactive() || identical(Sys.getenv("NOT_CRAN"), "true")) {
  dir.create("data-raw", showWarnings = FALSE, recursive = TRUE)
  if (!file.exists("./data-raw/rf_final_fit.rds")) {
    # rf_final_fit_bundled <- bundle(rf_final_fit)
    saveRDS(rf_final_fit, "./data-raw/rf_final_fit.rds")
  }
}

# rf_final_fit_bundled <- bundle(rf_final_fit)

# usethis::use_data(rf_final_fit, overwrite = TRUE, internal = TRUE)
```
